---
title: Multiple Regression
date: today
date-format: long
footer: "[DATS 1001 Website](https://ds4all.rocks)"
logo: images/ds4all-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: true
  message: false
  warning: false
  freeze: auto
---

## Load packages 

<br>

```{r} 
#| label: packages
#| echo: true

library(tidyverse)
library(vdemdata)
```


## Load VDEM Data {.smaller}

<br>

```{r}
#| label: get_data
#| output: false

model_data <- vdem |>
  filter(year == 2006) |> 
  select(country_name, 
         libdem = v2x_libdem, 
         wealth = e_gdppc, 
         oil_rents = e_total_oil_income_pc,
         polarization = v2cacamps, 
         corruption = v2x_corr, 
         judicial_review = v2jureview_ord, 
         region = e_regionpol_6C, 
         regime = v2x_regime) |> 
  mutate(
         region = factor(
           region,
           labels=c("Eastern Europe", 
             "Latin America", 
             "MENA", 
             "SSAfrica", 
             "Western Europe and North America", 
             "Asia and Pacific"))
          )

glimpse(model_data)
```

## Linear Model with Multiple Predictors

<br>

- Previously, we were interested in GDP per capita as a predictor of democracy
- Now, let's consider another predictor: polarization (also measured by V-Dem)

## Polarization Measure in the USA

```{r}
#| label: polarization_usa
#| echo: false

vdem |> 
  filter(country_name == "United States of America") |>
  filter(year > 1900) |> 
  ggplot(aes(y= v2cacamps, x = year)) +
  geom_line(color = "dodgerblue", size = 1) +
  labs(title = "Polarization in the USA", 
       y = "Polarization", x = "Year") +
  theme_minimal()
```

## Polarization and Democracy in 2006

```{r echo=FALSE}
ggplot(model_data, aes(x = polarization, y = libdem)) +
  geom_point() +
  geom_smooth(method = "lm", color = "#E48957", se = FALSE) +
  labs(x = "Polarization", y = "Liberal Democracy Index") +
  theme_minimal()
```

## Model with One Predictor

<br>

```{r}
#| label: model1

library(broom) # for tidy() function

lm(libdem ~ polarization, data = model_data) |> 
  tidy() # for nicer regression output
```
<br>

| How do we interpret the intercept and the slope estimate?
| Significance? 

## Model with Two Predictors

<br>

```{r}
#| label: model2
#| echo: true

lm(libdem ~ polarization + log(wealth), data = model_data) |> 
  tidy()
```

## Model with Two Predictors

<br>

$$ \hat{Y_i} = a + b_1*Polarization + b_2*GDPpc $$ 

$$ \hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc $$ 

## Model with Two Predictors 

<br>

$$ \hat{Y_i} = a + b_1*Polarization + b_2*GDPpc $$ 

$$ \hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc $$ 

$a$ is the predicted level of Y when BOTH GDP per capita and polarization are equal to 0

## Model with Two Predictors 

<br>

$$ \hat{Y_i} = a + b_1*Polarization + b_2*GDPpc $$ 

$$ \hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc $$ 

- $b_1$ is the impact of a 1-unit change in polarization on the predicted level of Y, **holding GDP per capita fixed** (all else equal)
- The relationship between polarization and democracy, controlling for wealth

## Model with Two Predictors 

$$ \hat{Y_i} = a + b_1*Polarization + b_2*GDPpc $$ 

$$ \hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc $$ 

- $b_2$ is the impact of a 1-unit change in GDP per capita on the predicted level of Y, **holding polarization fixed** (all else equal)
- The relationship between wealth and democracy, controlling for polarization

## Model with Two Predictors

<br>

$$ \hat{Y_i} = a + b_1*Polarization + b_2*GDPpc $$ 


- OLS is searching for combination of $a$, $b_1$, and $b_2$ that minimize sum of squared residuals
- Same logic as model with one predictor, just more complicated

## Model with Three Predictors 

<br>

$$ \hat{Y_i} = a + b_1*Polarization + b_2*GDPpc + b_3*OilRents $$ 

## Model with Three Predictors 

<br>

```{r}
#| label: model3
#| echo: true

lm(libdem ~ polarization + log(wealth) + oil_rents, data = model_data) |> 
  tidy()
```

## Model with Three Predictors {.smaller}

<br>

$$ \hat{Y_i} = a + b_1*Polarization + b_2*GDPpc + b_3*OilRents $$ 

$$ \hat{Y_i} = a + -.05*Polarization + .13*GDPpc + .00004*OilRents $$ 
<br>

| $b_3$ is the impact of a 1-unit change in oil revenues per capita on the predicted level of Y, **holding polarization and GDP fixed** (all else equal)

## Your Turn! {.smaller}

<br>

- In the last session, you examined levels of democracy and corruption
- Now, let's fit a multiple regression model predicting corruption with two predictors: democracy (libdem) and polarization (polarization)
- Then, interpret the coefficients
- Estimate a second multiple regression model that adds in GDP per capita (lg_gdppc)
- Interpret the coefficients
- What happens to the coefficients of the other variables when we add GDP per capita to the model?
    - Why do you think this happens?
- Try adding in additional predictors if there is time

```{r}
#| label: time1
#| echo: false

library(countdown)

countdown(minutes = 10, 
          id = "timer1", 
          bottom = "10%", 
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

## Model 1

<br>

```{r}
lm(corruption ~ libdem + polarization , data = model_data) |>
      tidy() 
```

## Model 2 - Adding GDP per capita

<br>

```{r}
lm(corruption ~ libdem + polarization + log(wealth), data = model_data) |>
      tidy() 
```

# Categorical Variables

## What is a Categorical Variable?

<br>

- A variable that represents distinct groups or categories
- Categories can be named (e.g., "Male" or "Female") 
- Or numbered (e.g., 1, 2, 3)
- Or words (like different regions)
- Can be either nominal (unordered) or ordinal (ordered)
- Also called "dummy variables" or "indicator variables"

## What is a Factor in R?

<br>

- A data structure used to represent categorical variables
- Can be dichotomous (binary) or have multiple levels
- Supports both ordered and unordered categories
- Call `factor()` to convert a variable to a factor
- Call `levels()` to see the categories

## Reference Categories

<br>

- In a regression model, every categorical variable needs a reference category
- The reference category is the category that all other categories are compared to when interpreting the results
- Call `relevel()` to change the reference category

## Judicial Review and Democracy

<br>

**Judicial Review:**

- Do high courts (Supreme Court, Constitutional Court, etc) have the power to rule on whether laws or policies are constitutional/legal? (Yes or No)
- Dimension of Judicial Independence

## Judicial Review and Democracy

```{r}
#| label: judicial_review
#| echo: false

ggplot(model_data, aes(x = wealth, y = libdem, 
                       color=factor(judicial_review))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "GPD per capita", y = "Liberal Democracy Index") +
  theme_bw() +
  scale_x_log10(labels = scales::dollar_format(suffix = "k")) +
  scale_color_manual(name = "Judicial Review", 
                     values = c("steelblue3", "coral"), 
                     labels = c("No", "Yes")) 

```

## Judicial Review and Democracy

<br>

```{r}
#| label: judicial_review_model

lm(libdem ~ factor(judicial_review), data = model_data) |>
  tidy()
```

## Judicial Review and Democracy {.smaller}

<br>

$$ \widehat{Democracy_{i}} = 0.17 + 0.28*JudicialReview(yes) $$

- **Slope:** countries with judicial review are expected, on average, to be 0.28 units more democratic on the liberal democracy index
  - Compares baseline level (Judicial Review = 0) to the other level (Judicial Review = 1)
- **Intercept:** average democracy score of countries without judicial review
- Average democracy score of countries with judicial review is 0.17 + 0.28 = 0.45

## Dummy Variables

<br>

- When the categorical explanatory variable has many levels, they're encoded to **dummy variables** 
- We always leave one category out of the model, as the omitted reference category
- Each coefficient describes is the expected difference between level of the factor **and the baseline level**
- Everything is relative to the omitted reference category

## Democracy and World Region

<br>

- Does region predict levels of democracy?
- Since Eastern Europe is the first category, default in R is to use that as the omitted category in models.

```{r}
#| label: region_model

levels(model_data$region)
```

## Democracy and World Region {.smaller}

<br>

How should we interpret intercept?  How about the coefficient on Latin America?

<br>

```{r}
#| label: region_model2

lm(libdem ~ region, data = model_data) |>
  tidy()
```

## Democracy and World Region {.smaller}

<br>

What if you want a different baseline category? How do we interpret now?

<br>


```{r}
#| label: region_model3

# make SS Africa the reference category
model_data <- model_data |> 
mutate(newReg = relevel(region, ref=4)) 

lm(libdem ~ newReg, data = model_data) |>
      tidy()
```

## Your Turn

<br>

**Which types of regime have more corruption?**

<br>
 
V-Dem also includes a categorial regime variable: Closed autocracy (0), Electoral Autocracy (1), Electoral Democracy (2), Liberal Democracy (3)

## Your Turn

<br>

**Which types of regime have more corruption?**

<br>

 
First, let's make this an easier factor variable to work with.

```{r}
#| label: regime_factor
#| echo: true

# Make nicer regime factor variable
model_data <- model_data |> 
  mutate(regime = factor(regime,
                         labels = c("Closed Autocracy",
                                   "Electoral Autocracy",
                                  "Electoral Democracy",
                                  "Liberal Democracy")))
levels(model_data$regime)
```

## Your Turn {.smaller}

<br>

**Which types of regime have more corruption?**

<br>

- Filter data to include only the year 2019 (or run the code to use model_data)
- Make a plot to visualize the relationship between regime type and level of corruption.          - Which kind of plot is best in this situation?
- Fit a linear model
- Interpret the intercept and coefficients


```{r}
#| label: time2
#| echo: false

library(countdown)

countdown(minutes = 10, 
          id = "timer2", 
          bottom = "10%", 
          right = "10%")
```

## Visualization

```{r}
#| label: corruption_plot
#| echo: false

model_data |> 
  group_by(regime) |> 
  summarize(corr = mean(corruption)) |> 
  ggplot(aes(y=corr, x = regime)) +
    geom_col(fill="steelblue") +
  theme_bw()
```

## Model

<br>

```{r}
#| label: corruption_model

lm(corruption ~ regime, data = model_data) |>
      tidy()
```

## Create Your Own Model {.smaller}

<br>

- What is a theory that you would like to test with V-Dem data?
- What is the dependent variable? 
- What are the independent variables? 
- Map out steps to wrangle the data and fit a regression model
- What do you expect to find? 
- Now go ahead and wrangle the data
- Fit the model
- Interpret the coefficients and their significance
- Did the results match your expectations?