---
title: Regression Practice
date: today
date-format: long
footer: "[DATS 1001 Website](https://ds4all.rocks)"
logo: images/ds4all-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

## Your Turn

<br>

**Are democracies less corrupt?**

<br>

- V-Dem includes a Political Corruption Index, which aggregates corruption in a number of spheres (see codebook for details).  

- The variable name is: *v2x_corr* : lower values mean less corruption


## Your Turn {.smaller}

<br>

**Are democracies less corrupt?**

<br>

::: {.smaller}
- Filter the V-Dem data to only include the year 2019
- Make a scatterplot to visualize the relationship between democracy (X) and corruption (Y) (use the *v2x_libdem* variable for democracy)
- Fit a linear model with `lm()`
- Interpret results for the slope and intercept
- For a country with the average (mean) level of democracy, what is the predicted level of corruption?
:::

```{r}
#| label: time1

library(countdown)

countdown(minutes = 15, 
          id = "timer1", 
          bottom = "5%", 
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

## Create Your Own Model {.smaller}

<br>

- What is a theory of democracy you could test with V-Dem?
- What is the dependent variable? 
- What is the independent variable? 
- Map out steps to wrangle the data and fit a regression model
- What do you expect to find? 
- Now go ahead and wrangle the data
- Fit the model
- Interpret the coefficients and their significance
- Did the results match your expectations?

```{r}
countdown(minutes = 15, 
          id = "timer2", 
          bottom = "5%", 
          right = "10%")
```


## Final Project 

<br>

- Project assignment 1 due this Sunday
- Can access [here](project/project-assignment-1.qmd)
- Let's get started!